\begin{thebibliography}{00}

\bibitem{9} M. Gambhir and V. Gupta, ``Recent automatic text summarization techniques: A survey,'' \textit{Artif. Intell. Rev.}, vol. 47, no. 1, pp. 1--66, 2021.
\bibitem{1} N. Liu and M. Lapata, ``Text summarization with pretrained encoders,'' in \textit{Proc. EMNLP}, 2019, pp. 1--15.
\bibitem{23} W. Li et al., "LayoutLM: Pre-training of Text and Layout for Document Understanding," in Proc. KDD, 2020.
\bibitem{10} P. Lopez, ``GROBID: Combining automatic bibliographic data recognition,'' \textit{Int. J. Digit. Libr.}, vol. 22, pp. 1--20, 2021.
\bibitem{3} J. Zhang et al., ``PEGASUS: Pre-training with extracted gap-sentences,'' in \textit{Proc. ICML}, 2020, pp. 1--12.
\bibitem{5} I. Beltagy et al., ``Longformer: The long-document transformer,'' \textit{arXiv:2004.05150}, 2020.
\bibitem{2} M. Lewis et al., ``BART: Denoising sequence-to-sequence pre-training,'' in \textit{Proc. ACL}, 2020, pp. 1--20.
\bibitem{4} C. Raffel et al., ``Exploring the limits of transfer learning with a unified T5,'' \textit{J. Mach. Learn. Res.}, vol. 21, pp. 1--67, 2020.
\bibitem{11} J. Devlin et al., ``BERT: Pre-training of deep bidirectional transformers,'' in \textit{Proc. NAACL-HLT}, 2019, pp. 1--15.
\bibitem{6} C. Y. Lin, ``ROUGE: A package for automatic evaluation of summaries,'' in \textit{Proc. ACL Workshop}, 2004, pp. 1--10.
\bibitem{12} A. Vaswani et al., ``Attention is all you need,'' in \textit{Proc. NeurIPS}, 2017, pp. 1--15.
\bibitem{16} S. Kulkarni et al., ``Towards building a scalable and low-latency summarization service,'' \textit{IEEE Cloud Computing}, vol. 10, no. 5, pp. 1--15, 2023.
\bibitem{8} J. Goldstein et al., ``Summarization via text extraction: Evaluation and corpus analysis,'' \textit{ACM Trans. Inf. Syst.}, vol. 18, no. 3, pp. 1--20, 2020.
\bibitem{13} R. Mehta et al., ``Web-scale NLP applications using Django and PyTorch,'' \textit{J. Web Eng.}, vol. 21, no. 4, pp. 1--25, 2022.
\bibitem{14} J. Maynez et al., ``On faithfulness and factuality in abstractive summarization,'' in \textit{Proc. EMNLP}, 2020, pp. 1--20.
\bibitem{15} F. Pedregosa et al., ``Scikit-learn: Machine learning in Python,'' \textit{J. Mach. Learn. Res.}, vol. 12, pp. 2825--2830, 2011.
\bibitem{7} R. Mihalcea and P. Tarau, ``TextRank: Bringing order into texts,'' in \textit{Proc. EMNLP}, 2004, pp. 1--10.
\bibitem{13} R. Mehta et al., ``Web-scale NLP applications using Django and PyTorch,'' \textit{J. Web Eng.}, vol. 21, no. 4, pp. 1--25, 2022.
\bibitem{14} J. Maynez et al., ``On faithfulness and factuality in abstractive summarization,'' in \textit{Proc. EMNLP}, 2020, pp. 1--20.
\bibitem{15} F. Pedregosa et al., ``Scikit-learn: Machine learning in Python,'' \textit{J. Mach. Learn. Res.}, vol. 12, pp. 2825--2830, 2011.
\bibitem{7} R. Mihalcea and P. Tarau, ``TextRank: Bringing order into texts,'' in \textit{Proc. EMNLP}, 2004, pp. 1--10.
\bibitem{24} T. Wolf et al., "Transformers: State-of-the-Art Natural Language Processing," in Proc. EMNLP, 2020.
\bibitem{19} A. Fabbri et al., ``SummEval: Re-evaluating summarization evaluation,'' \textit{Trans. Assoc. Comput. Linguist.}, vol. 9, pp. 1--20, 2021.
\bibitem{25} A. Kumar and L. Zhao, "NewsSumm: A Large-scale Dataset and Benchmarks for News Summarization," in Proc. ACL, 2021.
\bibitem{21} S. See et al., "Get To The Point: Summarization with Pointer-Generator Networks," in Proc. ACL, 2017.
\bibitem{22} Y. Liu, "Fine-tune BERT for Extractive Summarization," arXiv:1903.10318, 2021.
\bibitem{20} A. Roy et al., ``Interpretability and transparency in NLP: A survey,'' \textit{ACM Comput. Surv.}, vol. 57, no. 2, pp. 1--40, 2024.
\bibitem{26} S. Lee et al., "Efficient Transformers for Long Document Summarization," in Proc. NAACL, 2022.
\bibitem{27} M. Perez and J. Ortega, "Layout-aware Document Understanding with Graph Neural Networks," in Proc. ICDAR, 2021.
\bibitem{28} H. Chen et al., "Lightweight Summarization Models for Edge Deployment," IEEE Trans. Mobile Comput., vol. 22, no. 7, 2023.
\bibitem{29} Y. Sun and R. Gupta, "Hybrid Extractive-Abstractive Systems: A Survey and Evaluation," J. AI Research, vol. 68, pp. 123--156, 2022.
\bibitem{30} L. Fernandez et al., "Controlling Hallucinations in Abstractive Summarization," in Proc. EMNLP, 2022.
\bibitem{34} R. Banerjee and S. Iyer, "Benchmarking Factuality in Summarization," Trans. Assoc. Comput. Linguist., 2022.

\end{thebibliography}


